\documentclass{beamer}

\input{settings.tex}


\title{Semidefinite Programming}
\subtitle{\mycoursetitle, Lecture 8}
\author{by Sergei Savin}
\centering
\date{\mydate}



\begin{document}
\maketitle


%\begin{frame}{Content}
%
%\begin{itemize}
%\item  Semidefinite Programming (SDP)
%\item  Schur compliment
%\item  SOCP as SDP
%\item  Eigenvalues
%\item  Continuous Lyapunov equation as an SDP/LMI
%\item  Discrete Lyapunov equation as an SDP/LMI
%%\item  Example 3:  LMI Controller design for continuous LTI
%\end{itemize}
%\end{frame}



\begin{frame}{Semidefinite Programming (SDP)}
\begin{flushleft}

General form of a semidefinite program is:

%
\begin{equation}
\begin{aligned}
& \underset{\mathbf{x}}{\text{minimize}}
& & \mathbf{c}^\top\mathbf{x}, \\
& \text{subject to}
& & \begin{cases}
    \mathbf{G} + \sum\limits_{i=1}^n \mathbf{F}_i x_i \succeq 0, \\
    \mathbf{A}\mathbf{x} = \mathbf{b}.
    \end{cases}
\end{aligned}
\end{equation}

where $\mathbf{F}_i $ and $\mathbf{G}$ are symmetric, and $\succeq 0$ implies positive-semidefiniteness.

\bigskip

Constraint $\mathbf{G} + \sum\limits_{i=1}^n \mathbf{F}_i x_i \succeq 0$ is called \emph{linear matrix inequality} or \emph{LMI}.
 
\end{flushleft}
\end{frame}




\begin{frame}{SDP - Multiple LMI}
\begin{flushleft}

SDP can have several LMIs. Assume you have:

\begin{equation}
    \begin{cases}
        \mathbf{G} +  \mathbf{F} x_1 \succeq 0 \\
        \mathbf{H} x_2 \succeq 0
    \end{cases}
\end{equation}


This is equivalent to:

\begin{equation}
    \begin{bmatrix} 
           \mathbf{G} +  \mathbf{F} x_1 & \mathbf{0} \\
            \mathbf{0} & \mathbf{H} x_2 
    \end{bmatrix} 
    \succeq 0
\end{equation}

Symmatric block-diagonal matrix is positive semidefinite iff each of its blocks is positive semidefinite.

\end{flushleft}
\end{frame}



\begin{frame}{Trace of a matrix product}
	\begin{flushleft}
		
		Consider a matrices $\bo{E} = \begin{bmatrix}
			\bo{e}_1 & ... & \bo{e}_n
		\end{bmatrix}$ and 
		$\bo{X} = \begin{bmatrix}
			\bo{x}_1 & ... & \bo{x}_n
		\end{bmatrix}$. Their product can be written as:
		
		\begin{equation}
			\mathbf{E}\T \mathbf{X} = 
			\begin{bmatrix}
				\bo{e}_1 \T \bo{x}_1 & \bo{e}_1 \T \bo{x}_2 & ... \\
				\bo{e}_2 \T \bo{x}_1 & \bo{e}_2 \T \bo{x}_2 & ... \\
				... & ... & ... 
			\end{bmatrix}
		\end{equation}
		
		Thus, the trace of this product is given as:
		
		%
		\begin{equation}
			\text{tr}(\mathbf{E}\T\mathbf{X}) =  \bo{e}_1 \T \bo{x}_1 + ... + \bo{e}_n \T \bo{x}_n
		\end{equation}
		
		We can see that this is equivalent to an element-wise dot product. This represents a \emph{matrix inner product} (Frobenious inner product):
		
		%
		\begin{equation}
			\text{tr}(\mathbf{E}\T\mathbf{X}) =  \sum_{i,j} E_{ij} X_{ij}
		\end{equation}
		
		
		
	\end{flushleft}
\end{frame}



\begin{frame}{SDP matrix as a decision variable}
\begin{flushleft}

Sometimes it is easier use semidefinite matrices as decision variables directly:
%
\begin{equation}
\begin{aligned}
& \underset{\mathbf{X}}{\text{minimize}}
& & \text{tr}(\mathbf{E}\T \mathbf{X}), \\
& \text{subject to}
& & \begin{cases}
     \text{tr}(\mathbf{A}_i\T\mathbf{X}) = b_i, \\
     \mathcal{L}(\mathbf{X}) \succeq \mathbf{D}, \\
     \mathbf{X} \succeq 0
    \end{cases}
\end{aligned}
\end{equation}

where $ \mathcal{L}(\mathbf{X})$ is a linear function of $\mathbf{X}$.

\bigskip

In a cost function, matrix $\bo{E}$ plays the role of weights, similar to $\bo{f}$ in the linear cost $\bo{f}\T \bo{x}$. 

\end{flushleft}
\end{frame}






\begin{frame}{Continuous Lyapunov ineq. as SDP/LMI (1)}
	\begin{flushleft}
		
		In control theory, a sufficient condition for asymptotic stability of a continuous LTI system $\dot{\mathbf{x}} = \mathbf{A}\mathbf{x}$ is the existence of such $\mathbf{P} \succ 0$ that:
		
		\begin{equation}
			\mathbf{A}^\top\mathbf{P} + \mathbf{P}\mathbf{A}  \preceq -\mathbf{Q}
		\end{equation}
		%
		where $\mathbf{Q} \succ 0$ is a constant. This is called a \emph{Lyapunov criterion}. It can be represented as an SDP:
		%
		\begin{equation}
			\begin{aligned}
				& \underset{\mathbf{P}}{\text{minimize}}
				& & 0, \\
				& \text{subject to}
				& & \begin{cases}
					\mathbf{P} \succeq \epsilon \mathbf{I}, \\
					\mathbf{A}^\top\mathbf{P} + \mathbf{P}\mathbf{A} + \mathbf{Q} \preceq 0.
				\end{cases}
			\end{aligned}
		\end{equation}
		%
		where $\epsilon$ is a small positive constant.
		
	\end{flushleft}
\end{frame}




\begin{frame}{Continuous Lyapunov eq. as SDP/LMI (2)}
	\begin{flushleft}
		
		\input{code1}
		
	\end{flushleft}
\end{frame}




\begin{frame}{Discrete Lyapunov eq. as SDP/LMI (1)}
	\begin{flushleft}
		
		A discrete LTI system $\mathbf{x}_{i+1} = \mathbf{A}\mathbf{x}_i$ is asymptotically stable if there exists such $\mathbf{P} \succ 0$ that: 
		
		\begin{equation}
			\mathbf{A}^\top \mathbf{P}\mathbf{A} - \mathbf{P}  \preceq - \mathbf{Q}
		\end{equation}
		%
		where $\mathbf{Q} \succ 0$ is a constant. This can be represented as an SDP:
		%
		\begin{equation}
			\begin{aligned}
				& \underset{\mathbf{P}}{\text{minimize}}
				& & 0, \\
				& \text{subject to}
				& & \begin{cases}
					\mathbf{P} \succeq 0, \\
					\mathbf{A}^\top \mathbf{P}\mathbf{A} - \mathbf{P} + \mathbf{Q} \preceq 0.
				\end{cases}
			\end{aligned}
		\end{equation}
		
		
	\end{flushleft}
\end{frame}



\begin{frame}{Discrete Lyapunov eq. as SDP/LMI (2)}
	\begin{flushleft}
		
		\input{code2}
		
	\end{flushleft}
\end{frame}








\begin{frame}{Schur complement}
	\begin{flushleft}
		
		Schur complement. Given $\bo{M}$
		
		\begin{equation}
			\bo{M} = 
			\begin{bmatrix}
				\bo{A} & \bo{B} \\
				\bo{B}^\top & \bo{C}
			\end{bmatrix}
		\end{equation}
		
		with $\bo{A} \succ 0$, we can make the following statement:
		
		\begin{itemize}
			\item $\bo{M} \succ 0$ iff $\bo{C} - \bo{B}^\top\bo{A}^{-1}\bo{B} \succ 0$
		\end{itemize}
		
		\bigskip
		
		If $\bo{C}$ is full-rank, we can make the following statement:
		
		\begin{itemize}
			\item $\bo{M} \succ 0$ iff $\bo{C} \succ 0$ and $\bo{A} - \bo{B}\bo{C}^{-1}\bo{B}^\top \succ 0$
		\end{itemize}		
		
		
		
	\end{flushleft}
\end{frame}










\begin{frame}{SOCP as SDP}
	\begin{flushleft}
		
		Let us prove that SOCP is a subset of SDP. Affine SOC constraint is:
		
		\begin{equation}
			||\bo{A}\bo{x} + \bo{b}|| \leq \bo{c}^\top \bo{x} + d
		\end{equation}
		%
		where $\bo{c}^\top \bo{x} + d \geq 0$, and we can rewrite the SOC as: $(\bo{A}\bo{x} + \bo{b})^\top (\bo{A}\bo{x} + \bo{b}) \leq (\bo{c}^\top \bo{x} + d)^2$, and assuming $\bo{c}^\top \bo{x} + d > 0$ we can write it as:
		
		\begin{equation}
			\frac{(\bo{A}\bo{x} + \bo{b})^\top (\bo{A}\bo{x} + \bo{b})}{\bo{c}^\top \bo{x} + d} \leq \bo{c}^\top \bo{x} + d
		\end{equation}
		
		Transferring $\frac{(\bo{A}\bo{x} + \bo{b})^\top (\bo{A}\bo{x} + \bo{b})}{\bo{c}^\top \bo{x} + d}$ to the r.h.s:
		
		\begin{equation}
			-\frac{(\bo{A}\bo{x} + \bo{b})^\top (\bo{A}\bo{x} + \bo{b})}{(\bo{c}^\top \bo{x} + d)} + (\bo{c}^\top \bo{x} + d) \geq 0
		\end{equation}
		
	\end{flushleft}
\end{frame}





\begin{frame}{SOCP as SDP}
	\begin{flushleft}
		
		We can rewrite $-\frac{(\bo{A}\bo{x} + \bo{b})^\top (\bo{A}\bo{x} + \bo{b})}{(\bo{c}^\top \bo{x} + d)} + (\bo{c}^\top \bo{x} + d) \geq 0$ as:
		
		\begin{equation}
			(\bo{c}^\top \bo{x} + d) - 
			(\bo{A}\bo{x} + \bo{b})^\top 
			\bo{I}(\bo{c}^\top \bo{x} + d)^{-1}
			(\bo{A}\bo{x} + \bo{b})
			\geq 0
		\end{equation}
		
		\bigskip
		
		Using Schur we find an equivalent LMI:
		
		\begin{equation}
			\begin{bmatrix}
				\bo{I}(\bo{c}^\top \bo{x} + d) & (\bo{A}\bo{x} + \bo{b}) \\
				(\bo{A}\bo{x} + \bo{b})^\top & (\bo{c}^\top \bo{x} + d)
			\end{bmatrix} \succeq 0
		\end{equation}
		
		
	\end{flushleft}
\end{frame}



\begin{frame}{Matrix variable in affine SOC}
	\begin{flushleft}
		
		Consider the following constraint, where $\bo{X}$ is a variable:
		
		\begin{equation}
			\label{eq:norm_LMI}
			||\bo{X}\bo{v} + \bo{b}|| \leq \bo{c}^\top \bo{y} + d
		\end{equation}
		
		Can we re-write it as an LMI? Using the same process as before we get:
		
		\begin{align}
			\begin{bmatrix}
	(\bo{c}^\top \bo{y} + d) & (\bo{X}\bo{v} + \bo{b}) \\
	(\bo{X}\bo{v} + \bo{b})^\top & (\bo{c}^\top \bo{y} + d)
		\end{bmatrix} \succeq 0
		\end{align}
	
		So, \eqref{eq:norm_LMI} is an admissible constraint in an SDP.
		
	\end{flushleft}
\end{frame}






\begin{frame}{SDP and eigenvalues}
	\begin{flushleft}
		
		Consider the problem: find the argest eigenvalue of $A$. The solution is:
		
		\begin{equation}
			\begin{aligned}
				& \underset{t}{\text{minimize}}
				& & t, \\
				& \text{subject to}
				& & \bo{A} \preceq t\bo{I}
			\end{aligned}
		\end{equation}
		
		Proof. If $\lambda$ is an eigenvalue of $\bo{A}$, then $\bo{A} \bo{v} = \lambda\bo{v}$, hence $(\bo{A} - t\bo{I}) \bo{v} = (\lambda - t)\bo{v}$, meaning $\lambda - t$ is eigenvalue of $(\bo{A} - t\bo{I})$. Thus, if $(\bo{A} - t\bo{I})$ is negative semi-definite, then $\lambda - t \leq 0$ and $\lambda \leq t$. \qed
		
		
	\end{flushleft}
\end{frame}





\begin{frame}{Homework}
\begin{flushleft}



Implement both examples from page 2 of the LMI CVX documents: \bref{http://stanford.edu/class/ee363/notes/lmi-cvx.pdf}{stanford.edu/class/ee363/notes/lmi-cvx.pdf}.

\end{flushleft}
\end{frame}




\begin{frame}{Literature}
	\begin{flushleft}
		
		\begin{itemize}
			\item Introduction to Semidefinite Programming (SDP), Robert M. Freund \bref{https://ocw.mit.edu/courses/15-084j-nonlinear-programming-spring-2004/a632b565602fd2eb3be574c537eea095\_lec23\_semidef\_opt.pdf}{https://ocw.mit.edu/courses/15-084j-nonlinear-programming-spring-2004/}
			
			\item Semidefinite Programming, L. Vandenberghe and S. Boyd: \bref{https://stanford.edu/~boyd/papers/pdf/semidef\_prog.pdf}{https://stanford.edu/~boyd/papers/pdf/semidef\_prog.pdf}. Main page: \bref{https://stanford.edu/~boyd/papers/sdp.html}{https://stanford.edu/~boyd/papers/sdp.html}
		\end{itemize}
		
		
		
	\end{flushleft}
\end{frame}



\myqrframe



\begin{frame}{Schur Complement (Quadratic Form Proof), 1}
	\begin{flushleft}
		
		Let 
		\[
		\bo{M} = 
		\begin{bmatrix}
			\bo{A} & \bo{B} \\
			\bo{B}^\top & \bo{C}
		\end{bmatrix},
		\quad \bo{A} = \bo{A}^\top \succ 0.
		\]
		
		We prove:
		\[
		\bo{M} \succ 0 
		\quad \Longleftrightarrow \quad
		\bo{C} - \bo{B}^\top \bo{A}^{-1} \bo{B} \succ 0.
		\]
		
		\bigskip
		
		Consider the quadratic form:
		\begin{align}
			f &= 
			\begin{bmatrix}
				\bo{x}^\top & \bo{y}^\top
			\end{bmatrix}
			\begin{bmatrix}
				\bo{A} & \bo{B} \\
				\bo{B}^\top & \bo{C}
			\end{bmatrix}
			\begin{bmatrix}
				\bo{x} \\ \bo{y}
			\end{bmatrix} \\
			&= \bo{x}^\top \bo{A} \bo{x} + 2 \bo{x}^\top \bo{B} \bo{y} + \bo{y}^\top \bo{C} \bo{y}.
		\end{align}
		
		\bigskip
		
		Complete the square:
		\begin{align*}
			f &= 
			(\bo{x} + \bo{A}^{-1}\bo{B}\bo{y})^\top 
			\bo{A} 
			(\bo{x} + \bo{A}^{-1}\bo{B}\bo{y})
			+ \bo{y}^\top (\bo{C} - \bo{B}^\top \bo{A}^{-1}\bo{B}) \bo{y}.
		\end{align*}
		
	\end{flushleft}
\end{frame}



\begin{frame}{Schur Complement (Quadratic Form Proof), 2}
	\begin{flushleft}
		%
		{\small
		\textcolor{mygray}{
		\[
			f = 
			(\bo{x} + \bo{A}^{-1}\bo{B}\bo{y})^\top 
			\bo{A} 
			(\bo{x} + \bo{A}^{-1}\bo{B}\bo{y})
			+ \bo{y}^\top (\bo{C} - \bo{B}^\top \bo{A}^{-1}\bo{B}) \bo{y}.
		\]
		}
		%
		\textbf{($\Rightarrow$)} Suppose $\bo{M} \succ 0$.
		%
		\begin{itemize}
			\item Let $\bo{y} = 0$:
			\[
			f = \bo{x}^\top \bo{A} \bo{x} > 0 \;\Rightarrow\; \bo{A} \succ 0.
			\]
			
			\item Let $\bo{x} = -\bo{A}^{-1}\bo{B}\bo{y}$:
			\[
			f = \bo{y}^\top (\bo{C} - \bo{B}^\top \bo{A}^{-1}\bo{B}) \bo{y} > 0
			\]
			\[
			\Rightarrow\; \bo{C} - \bo{B}^\top \bo{A}^{-1}\bo{B} \succ 0.
			\]
		\end{itemize}
		%
		\textbf{($\Leftarrow$)} Suppose 
		\[
		\bo{A} \succ 0, \quad 
		\bo{C} - \bo{B}^\top \bo{A}^{-1}\bo{B} \succ 0.
		\]
		
		Then both terms in $f$ are positive for all $(\bo{x},\bo{y}) \neq 0$, hence:
		\[
		f > 0 \;\Rightarrow\; \bo{M} \succ 0.
		\]
		%
		\textbf{Conclusion:}
		\[
		\bo{M} \succ 0 
		\;\Longleftrightarrow\;
		\bo{A} \succ 0 \;\text{and}\;
		\bo{C} - \bo{B}^\top \bo{A}^{-1}\bo{B} \succ 0.
		\]}
		
	\end{flushleft}
\end{frame}



\end{document}
